---
title: "Prostate phosphoSWATH: Enrichment CARNIPHAL Results"
author: "Alberto Valdeolivas: alberto.valdeolivas@bioquant.uni-heidelberg.de; Date:"
date: "18/02/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

Study the phospho-proteome of two prostate cancer cell lines upon perturbation 
with a combination of different ligands and inhibitors. This vignette is focused
on generating cluster of phosphosites following a given trend in our data. 

## Reading and Formatting linear model results

We first loaded the required libraries for the forthcoming analysis.

```{r, message=FALSE}
library(OmnipathR)
library(dplyr)
library(readr)
library(igraph)
library(bc3net)
library(ggplot2)
library(gridExtra)
```

Then, we read and format the results of our linear model. 

```{r, message=FALSE}
ResultsLinearModel <- read_tsv("Data/limma_model_results_20190301.tsv")

LinearModelData_df <- ResultsLinearModel %>% 
    dplyr::filter(!is.na(residues_str)) %>% 
    dplyr::mutate(residues_str = strsplit(residues_str, "_")) %>% 
    tidyr::unnest(residues_str) %>% 
    dplyr::mutate(GeneSymbol_Residue = paste(GeneSymbol, residues_str, sep="_")) 

covariate_vars <- c("Intercept","ms_day180409", "ms_day180410", "ms_day180412", 
                    "ms_day180414", "Culture_batch", "fraction_missing", "LNCaP")

## From linear model to Matrix suitable to run Viper.
LinearModelData_df_Clean <-     
    LinearModelData_df %>% 
    dplyr::select(statistic,GeneSymbol_Residue, term, p.value) %>%
    dplyr::filter(!(term %in% covariate_vars)) %>%
    dplyr::filter(GeneSymbol_Residue != "NA_SNA") %>%
    dplyr::filter(GeneSymbol_Residue != "NA_TNA") %>%
    dplyr::group_by(GeneSymbol_Residue, term) %>%
    dplyr::filter(p.value == min(p.value)) %>%
    dplyr::ungroup() %>%
    dplyr::distinct() %>% 
    dplyr::select(-p.value)  
```

We focus on the 500 most variable phosphosites to avoid finding
cluster of sites that do not vary under these experimental conditions. 

```{r, message=FALSE}
MostVariableSites <- LinearModelData_df_Clean %>% 
    dplyr::group_by(GeneSymbol_Residue) %>%
    dplyr::mutate(variance = var(statistic))  %>%
    dplyr::arrange(desc(variance)) %>%
    dplyr::distinct(GeneSymbol_Residue, .keep_all = TRUE) %>%
    dplyr::ungroup() %>%
    dplyr::top_n(500) %>%
    dplyr::select(GeneSymbol_Residue)
```

## Computing correlations and Finding Communities of phosphosites

We generate a matrix with all the phosphosites and the conditions and we 
compute Pearson correlation on it. Then, we create a network of phosphosites by
drawing an edge between a pair of phosphosites if they have a correlation larger
than 0.7. This is an initial selection that can be improved or modified. We
also create an anticorrelation graph, although we will now move on with the 
positive correlation graph. 

```{r, message=FALSE}
MatrixStatistic <- 
    dplyr::semi_join(LinearModelData_df_Clean,MostVariableSites) %>%
    tidyr::pivot_wider(names_from = term, values_from = statistic)  %>%
    tibble::column_to_rownames(var = "GeneSymbol_Residue") %>% 
    as.matrix()

corMatrix <- 
    cor(t(MatrixStatistic), use = "pairwise.complete.obs",method = c("pearson"))

corGraph_AdjMatrix <- 
    apply(corMatrix, 2, function(x){ ifelse(x >= 0.7,1,0)})
AnticorGraph_AdjMatrix <- 
    apply(corMatrix, 2, function(x){ ifelse(x <= -0.7,1,0)})    
```

Then, we apply Louvain clustering on the graph. It is to note that we apply an
iterative version of Louvain. The method is applied in a recursive way for those 
communities larger than a given size. We chose communities containing between 5
and 40 phosphosites. This election is arbitrary but based on the fact that a 
cluster of this number of phosphosites is easier to interpret from a biological
perspective (check how many phosphosites are in avergare regulated by a kinase).

```{r, message=FALSE}
## Transform to an igraph object
corGraph_igraph <- graph_from_adjacency_matrix(corGraph_AdjMatrix, 
    mode = c("undirected"), diag = FALSE)

## I get the largest connected component of the graph: 
corGraph_igraph <- getgcc(corGraph_igraph)

## Function to Perform an iterative Louvain Clustering.
IterativeLouvain <- function(x,OriginalNetwork,minSize,maxSize) {
  a <- 0 
  n <- length(x)
  newList <- list()
  for (i in seq(n)){
    currentGenes <- x[[i]]
    if (length(currentGenes) >= minSize){
      if(length(currentGenes) <= maxSize){
        a <- a + 1 
        newList[a]  <-x[i]
      } else {
        Subnetwork <-induced_subgraph(OriginalNetwork, currentGenes)
        Modules_rec <- cluster_louvain(Subnetwork, weights = E(Subnetwork)$weight)  
        m <- length(Modules_rec)
        for (j in seq(m)){
          if (length(Modules_rec[[j]]) >= minSize ){
            a <- a + 1 
            newList[[a]] <- Modules_rec[[j]]
          }
        }  
      } 
    }
  }
  return(newList)
}

## We define the size of the modules and the maximun number of iterations. 
smallest_size <- 5
largest_size <- 40
iter <- 0

## First Louvain clustering to obtain the first modules
Modules <- cluster_louvain(corGraph_igraph)

## Iterative clusters to split the first modules into smaller clusters.
while (max(sapply(Modules[],length)) >= largest_size && iter <= 100) {
  iter <-iter +1 
  # print (iter)
  NewModules <- IterativeLouvain(Modules,corGraph_igraph,smallest_size,largest_size)
  Modules <- NewModules  
}
```

## Plotting the Results 

In this case, we focus on analysing the trend of phospohsites in the 
different time points between the two cell lines when no inhibition is applied
and the EGF ligand is present. 

We first select the experimental conditions of interest and the phosphosites of
a given module and extract their t-values from our linear model. 

```{r, message=FALSE}
## I select my conditions of interest
ConditionA <- 
    c("LNCaP_noInhib_t0_noLigand", "LNCaP_noInhib_t1_EGF","LNCaP_noInhib_t2_EGF")
ConditionB <- 
    c("abl_noInhib_t0_noLigand", "abl_noInhib_t1_EGF","abl_noInhib_t2_EGF")

## I take as an one of our modules:
SitesModule <- Modules[[21]]
SitesModule

## We extract the info for these conditions
LinearModelData_df_SelectA <- LinearModelData_df_Clean %>%
    dplyr::filter(term %in% ConditionA, GeneSymbol_Residue %in% SitesModule) %>%
    dplyr::mutate(time = ifelse(grepl("_t0_", term), "t0", 
                         ifelse(grepl("_t1_", term), "t1","t2")))

LinearModelData_df_SelectB <- LinearModelData_df_Clean %>%
    dplyr::filter(term %in% ConditionB, GeneSymbol_Residue %in% SitesModule) %>%
    dplyr::mutate(time = ifelse(grepl("_t0_", term), "t0", 
                         ifelse(grepl("_t1_", term), "t1","t2")))
```

We finally plot the results.

```{r, message=FALSE}
pA <- ggplot(data=LinearModelData_df_SelectA, aes(x=time, y=statistic, group=GeneSymbol_Residue)) +
    geom_line(alpha = 0.5, color = "blue", linetype = "dashed", size= 1.5) +
    geom_point(alpha = 0.5, size = 3,  color = "blue") +  
    theme_minimal() + ggtitle("LNCaP + EGF") + 
    theme(plot.title = element_text(size = 20, face = "bold", colour = "red", 
                                    hjust = 0.5))


pB <- ggplot(data=LinearModelData_df_SelectB, aes(x=time, y=statistic, group=GeneSymbol_Residue)) +
    geom_line(alpha = 0.5, color = "blue", linetype = "dashed", size= 1.5) +
    geom_point(alpha = 0.5, size = 3,  color = "blue") +  
    theme_minimal() + ggtitle("abl + EGF")   + 
    theme(plot.title = element_text(size = 20, face = "bold", colour = "red",
                                    hjust = 0.5))

grid.arrange(pA, pB, ncol=2)
```

## Conclusions

This is just an illustration of what we aim at finding clustering the 
phosphosites. However,it is to note that the procedure described in this 
vignette is not the most optimal. We computed the correlations taking into
account all the experimental conditions and later we only focused on a given
one. Therefore, some phosphosites in the cluster of the analysed condition
can have an unexpected behaviour. We can define a better procedure based on 
what we want to analyse. I think that it would be very interesting to find 
clusters of phosphosites that have a similar behaviour/trend in one condition
and the opposite in another. 


## R session Info

```{r}
sessionInfo()
```
